---
title: "Fitting 2D Gaussians to Data"
author: "Vikram B. Baliga"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting 2D Gaussians to Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{metR}
  %\VignetteDepends{lattice}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The function `fit_gaussian_2D()` can be used fit 2D Gaussians to data, and has
several methods for how the fitting is implemented. This vignette will run you
through what these methods mean with worked examples.

We'll begin by loading `gaussplotR` and loading the sample data set provided
within. The raw data we'd like to use are in columns 1:3, so we'll shave the
data set down to those columns before running through the examples.

```{r setup}
library(gaussplotR)

## We'll also use lattice, ggplot2 and metR
library(lattice); library(ggplot2); library(metR)

## Load the sample data set
data(gaussplot_sample_data)

## The raw data we'd like to use are in columns 1:3
samp_dat <-
  gaussplot_sample_data[,1:3]
```

It generally helps to plot the data beforehand to get a sense of its overall 
shape. We'll simply produce a contour plot.

```{r raw_data_contour}
lattice::levelplot(
  response ~ X_values * Y_values,
  data = samp_dat,
  col.regions = colorRampPalette(
    c("white", "blue")
    )(100),
  xlim = c(-5, 0),
  ylim = c(-1, 4), 
  asp = 1
)

```


## The `method` and `constrain_orientation` arguments

### `method`

`gaussplotR::fit_gaussian_2D()` has three main options for its `method` 
argument: 1) `"elliptical"`, 2) `"elliptical_log"`, or 3) `"circular"`. 

The most generic method (and the default) is `method = "elliptical"`. This
allows the fitted 2D Gaussian to take an ellipsoid shape. If you would like the
best-fitting 2D Gaussian, this is most likely your best bet.

A slightly-altered method to fit an ellipsoid Gaussian is available in 
`method = "elliptical_log"`. This method follows Priebe et al. 2003^[Priebe NJ, 
Cassanello CR, Lisberger SG. The neural representation of speed in macaque area 
MT/V5. J Neurosci. 2003 Jul 2;23(13):5650-61. doi: 
10.1523/JNEUROSCI.23-13-05650.2003.] and is geared towards use with 
log2-transformed data.

A third option is `method = "circular"`. This produces a very simple 2D Gaussian
that is constrained to have to have a roughly circular shape (i.e. spread in X-
and Y- are roughly equal).

An additional argument, `constrain_orientation` gives additional control over the
orientation of the fitted Gaussian.  By default, the `constrain_orientation` is
`"unconstrained"`, meaning that the best-fit orientation is returned. 

### `constrain_orientation`
Setting constrain_orientation to a numeric (e.g. `constrain_orientation = pi/2`)
will force the orientation of the Gaussian to the specified value, but this is
only available when using `method = "elliptical"` or `method = "elliptical_log"`

Note that supplying a numeric to `constrain_orientation ` is handled differently 
by `method = "elliptical"` vs `method = "elliptical_log"`. With 
`method = "elliptical"`, a `theta` parameter dictates the rotation, in 
radians, from the x-axis in the clockwise direction. Thus, using 
`method = "elliptical", constrain_orientation = pi/2` will return parameters for
an elliptical 2D Gaussian that is constrained to a 90-degree (pi/2) orientation. 
In contrast, the `method = "elliptical_log"` procedure uses a `Q` parameter to 
determine the orientation of the 2D Gaussian. 
Setting `method = "elliptical_log", constrain_orientation = 0` will result 
in a diagonally-oriented Gaussian, whereas setting `constrain_orientation = -1`
will result in horizontal orientation. Again, see Priebe et al. 2003 for more 
details.

## Example 1: Unconstrained elliptical

Unconstrained ellipticals are the default option and are generally recommended
for most purposes. Here's an example:

```{r u_e}
gauss_fit_ue <-
    fit_gaussian_2D(samp_dat)

gauss_fit_ue
attributes(gauss_fit_ue)
```
Fitting an unconstrained ellipse returns an object (here: `gauss_fit_ue`) that
is a `data.frame` with one column per fitted parameter. The fitted parameters
are: `A_o` (a constant term), `Amp` (amplitude), `theta` (rotation, in radians,
from the x-axis in the clockwise direction), `X_peak` (x-axis peak location),
`Y_peak` (y-axis peak location), `a` (width of Gaussian along x-axis), and `b`
(width of Gaussian along y-axis).

Note that the attribute `fit_method` indicates that an `"elliptical_unconstr"`
approach was used. This attribute is used by `predict_gaussian_2D()` to 
automatically determine what method (and therefore, identity of parameters) was 
used and then sample points from that fitted Gaussian. 

We can elect to sample more points from the fitted Gaussian by feeding in a
grid of x- and y- values on which to predict (via `expand.grid()`. 

Then, the fitted object `gauss_fit_ue` along with the grid of points can be 
supplied to `predict_gaussian_2D` to sample more points from the fit, which
can be useful for plotting.

```{r predict_and_plot_ue}
## Generate a grid of x- and y- values on which to predict
grid <-
  expand.grid(X_values = seq(from = -5, to = 0, by = 0.1),
              Y_values = seq(from = -1, to = 4, by = 0.1))

## Predict the values using predict_gaussian_2D
gauss_data_ue <-
  predict_gaussian_2D(
    fit_object = gauss_fit_ue,
    X_values = grid$X_values,
    Y_values = grid$Y_values,
  )

## Plot via ggplot2 and metR
ggplot_gaussian_2D(gauss_data_ue)
```

## Example 2: Constrained elliptical

As noted above, the `constrain_orientation` can be used to dictate the 
orientation. Please note that this will very likely result in a poorer fit, but
may be useful for certain types of analyses.

Here we'll force the Gaussian to be horizontally-oriented. 

```{r c_e}
gauss_fit_ce <-
    fit_gaussian_2D(samp_dat, 
                    constrain_orientation = 0)

gauss_fit_ce
```

We'll use the same grid of x- and y- points as above

```{r predict_and_plot_ce}
## Predict the values using predict_gaussian_2D
gauss_data_ce <-
  predict_gaussian_2D(
    fit_object = gauss_fit_ce,
    X_values = grid$X_values,
    Y_values = grid$Y_values,
  )

## Plot via ggplot2 and metR
ggplot_gaussian_2D(gauss_data_ce)
```

## Example 3: Unconstrained elliptical_log

This procedure follows the formula used in Priebe et al. 2003 and is geared 
towards log2-transformed data (which the example data are).

Parameters from this model include: `Amp` (amplitude), `Q` (orientation
parameter), `X_peak` (x-axis peak location), `Y_peak` (y-axis peak location),
`X_sig` (spread along x-axis), and `Y_sig` (spread along y-axis).

```{r uel}
gauss_fit_uel <-
    fit_gaussian_2D(samp_dat, 
                    method = "elliptical_log")

gauss_fit_uel

## Predict the values using predict_gaussian_2D
gauss_data_uel <-
  predict_gaussian_2D(
    fit_object = gauss_fit_uel,
    X_values = grid$X_values,
    Y_values = grid$Y_values,
  )

## Plot via ggplot2 and metR
ggplot_gaussian_2D(gauss_data_uel)
```

## Example 4: Constrained elliptical_log

Similar to the above, but here the `constrain_orientation` can be used to dictate
the value of the `Q` parameter used in Priebe et al. 2003. Setting `Q` to 0 will
result in a diagonally-oriented Gaussian, whereas setting `Q` to -1 will result
in horizontal orientation. `Q` is a continuous parameter, so values in between
may be used as well, such as in this example:

```{r cel}
gauss_fit_cel <-
    fit_gaussian_2D(
      samp_dat,
      method = "elliptical_log",
      constrain_orientation = -0.66
    )

gauss_fit_cel

## Predict the values using predict_gaussian_2D
gauss_data_cel <-
  predict_gaussian_2D(
    fit_object = gauss_fit_cel,
    X_values = grid$X_values,
    Y_values = grid$Y_values,
  )

## Plot via ggplot2 and metR
ggplot_gaussian_2D(gauss_data_cel)
```

Again, setting the value of `Q` via `constrain_orientation` will very likely 
result in poorer-fitting Gaussians. See the analyses in Priebe et al. 2003 to
get a sense of useful applications of this approach. Forcing Q = -0.66 in 
the above example isn't all that useful, but goes to show that it can be done.

## Example 5: Circular

Using `method = "circular"` constrains the Gaussian to have a roughly circular
shape (i.e. spread in X- and Y- are roughly equal).

If this method is used, the fitted parameters are: `Amp` (amplitude), `X_peak`
(x-axis peak location), `Y_peak` (y-axis peak location), `X_sig` (spread along
x-axis), and `Y_sig `(spread along y-axis).

```{r cir}
gauss_fit_cir <-
    fit_gaussian_2D(samp_dat, 
                    method = "circular")

gauss_fit_cir

## Predict the values using predict_gaussian_2D
gauss_data_cir <-
  predict_gaussian_2D(
    fit_object = gauss_fit_cir,
    X_values = grid$X_values,
    Y_values = grid$Y_values,
  )

## Plot via ggplot2 and metR
ggplot_gaussian_2D(gauss_data_cir)
```

That's all!  

ðŸ¢
